Using Metric Space Indexing for Complete and Efficient Record Linkage

### 2017-11-21 paper.tex
Update on Overleaf.

Record linkage is the process of identifying records that refer to the
same real-world entities, in situations where entity identifiers are
unavailable. Records are linked on the basis of similarity between
common attributes, with every pair of records being classified as a
link or non-link depending on the degree of similarity between them.
Record linkage is usually performed in a three-step process: first
groups of similar candidate records are identified using indexing; 
pairs within the same group are then compared in more detail, and
finally classified.
%
Even state-of-the-art indexing techniques, such as Locality Sensitive
Hashing, have potential drawbacks. They may fail to group together some
true matching records with high similarity. Conversely, they may group
records with low similarity, leading to high computational overhead.
%
We propose using metric space indexing to perform \emph{complete} record
linkage, which results in a parameter-free record linkage process
combining indexing, comparison and classification into a single step
delivering complete and efficient record linkage.

@@

To overcome data quality issues such as typographical errors and
variations (which are common in name and address values~\cite{Chr12}),
approximate string comparison functions (such as edit distance, the
Jaro-Winkler comparator, or Jaccard similarity~\cite{Chr12}) are used to
compare pairs of records, leading to a vector of similarities (one
similarity per attribute/field compared) for each pair. These similarity
vectors are then used to classify the compared record pairs into links
(where it is assumed both records in a pair correspond to the same
real-world entity) and non-links (where the records are assumed to
correspond to two different entities). Various classification methods
have been employed in record linkage~\cite{Chr12,Don15}, ranging from
simple threshold-based to sophisticated clustering, supervised
classification techniques, and active learning approaches~\cite{Wan15}.

@@

Systems that perform indexing prior to comparison and classification, as
illustrated in Fig.~\ref{fig-rl-process}, add a further practical
complexity to the process. Indexing, comparison and classification are
often conducted using algorithms and parameters selected based on
technical and domain expertise of the user of the record linkage system,
followed by a manual assessment (auditing) of the linkage
outcomes~\cite{Chr12}. If the quality of the resulting links is not good
enough for a certain application, the linkage process needs to be
repeated with different parameter settings and possibly also alternative
algorithms. This can result in a time-consuming and labour-intensive
iterative process. A major challenge is often the heuristic nature of
indexing, where the choice of a certain indexing technique and its
parameters (including which attributes to use in indexing) will
determine the final outcome of a linkage.

This paper focuses on approaches that use a similarity threshold to
classify links. Such techniques are fundamentally limited by the extent
to which true matching records are similar, and true non-matches are
dissimilar---this is dataset-dependent. Within this domain, we define
a technique to be \emph{complete} if it is guaranteed to identify all
pairs within the specified similarity threshold. Commonly used indexing
techniques are incomplete, since they reduce computational cost at the
expense of potentially overlooking some true matches. By definition,
incomplete techniques yield lower recall than complete techniques.
Conversely, and perhaps counter-intuitively, complete techniques can
yield lower precision with some datasets. This is discussed further in
Sect.~\ref{sec-approach}.

\emph{Metric space indexing} (MSI) is a complete technique with much
lower computational overheads than a brute force approach. It allows
indexing, comparison and classification to be combined into a single
step, as shown in Fig.~\ref{fig-rl-process}, making the overall record
linkage process simpler, more efficient and more effective than
incomplete indexing approaches.

\textbf{Contribution:} the primary contribution of this paper is the
novel application of MSI to achieve complete and efficient record
linkage, without the need for complex parameter tuning. We evaluate our
approach on several datasets from diverse domains and demonstrate its
advantages over existing indexing techniques for record linkage.

@@

A different approach to indexing is the sorted neighbourhood
method~\cite{Mon96}, where the datasets to be linked are sorted
according to a \emph{sorting key} (usually a concatenation of the values
from several attributes). A sliding window is then moved over the
datasets and only records within the window are compared. Techniques
that adaptively shrink or expand the window size based on the
characteristics of the sorting key values have shown to improve both
linkage efficiency and quality~\cite{Dra12} over other sliding window
approaches.

@@

Locality sensitive hashing (LSH), originally proposed to allow efficient
nearest-neighbour search in high-dimensional spaces~\cite{Ind98}, has
been employed in record linkage as an indexing technique where attribute
values are hashed multiple times, and blocks are created from those
records that share some hash values. \emph{HARRA}~\cite{Kim10} is a
record linkage approach based on MinHash~\cite{Broder1997} and LSH which
blocks, compares, and then merges linked records in an iterative
fashion, where merged records are re-hashed to improve overall linkage
quality. ~\cite{Steorts2014} evaluates two LSH variations, concluding
that in order to get good results, LSH methods must be tuned to the
particular datasets being linked. This requires good quality ground
truth data which may be unavailable or expensive to obtain.

Metric space indexing (MSI) techniques provide indexing structures to
support the comparison of records in one set with those in another,
typically also offering similarity search operations. To create an MSI
data structure, it is necessary to define a distance measure between
records, with certain properties including the \emph{triangle
inequality}. Similarity search operations include
\textit{range-search(q,d)}, where all records within a distance $d$ of a
query record $q$ are identified; \textit{nearest-neighbour(q)},
returning the record with smallest distance to $q$; and
\textit{nearest-n(q,n)}, returning the $n$ closest records to $q$.
\cite{Zezula2010} gives an overview of MSI techniques.
Here we choose one metric space indexing structure, the M-tree
\cite{paolociaccia2m}, and investigate its efficacy for record linkage.

@@

\cite{Li2006} describes a threshold-based linkage method using
R-trees~\cite{Hjaltason1998}. It demonstrates that high quality may be
achieved using Jaccard similarity over selected record fields.
\cite{Ciaccia97indexingmetric} shows that M-trees are almost always more
efficient than R-trees, hence their use in the experiments described in
this paper.

@@

\Cref{cora-quality} shows the precision, recall, and F-measure
for various thresholds. As expected, low thresholds give high precision and low recall, and the reverse for high threshold
values. As expected, brute force and M-tree give identical results.

For a more detailed investigation of selected linkers, the brute force
approach is used to establish the optimum threshold value for the Cora
dataset. The maximum F-measure value is observed at a threshold value of
70. This value is dataset-dependent; for different datasets the maximum
F-measure will occur at different thresholds. In the rest of this
section we fix the threshold value at 70.

\Cref{comparison-of-results-cora} shows greater detail for selected
linkers, with the distance threshold fixed at 70. The table shows the
parameters for the experiment, the number of distance comparisons made,
the precision, recall and the F1 measure achieved by each algorithm.  In
the \emph{Linker} column the algorithm name is followed by its
parameters: for LSH the number of the bands followed by the band size,
and for traditional blocking the fields used for blocking. The number of
distance comparisons is reported as a machine-independent proxy for
execution cost, since code profiling shows that distance calculations
are dominant.

@@

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-brute}
\caption{Brute Force}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-mtree}
\caption{M-tree}
\end{subfigure}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-authors}
\caption{Blocking on `authors' field}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-title}
\caption{Blocking on `title' field}
\end{subfigure}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-year}
\caption{Blocking on `publication year' field}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-combined}
\caption{Blocking using the union of all fields}
\end{subfigure}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-lsh-2-2}
\caption{LSH using 2 bands of size 2}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-lsh-2-10}
\caption{LSH using 2 bands of size 10}
\end{subfigure}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-lsh-10-2}
\caption{LSH using 10 bands of size 2}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-lsh-10-10}
\caption{LSH using 10 bands of size 10}
\end{subfigure}

@@

\Cref{comparison-of-results-demography-skye} and \Cref{comparison-of-results-demography-kili} show the results for LSH-MinHash and M-tree on the Skye and Kilmarnock datasets. Linkage was performed between birth and death records. The shingle size was set to 2 for all the LSH experiments reported, as this was found to give good results and LSH was not especially sensitive to this parameter. Results for other shingle sizes were omitted from this paper for brevity. In both cases the F1 measure achieved is better for M-tree than any of the experimental settings chosen for LSH. This is largely due to the recall for M-tree being much higher than any of the LSH experiments. For most configurations, LSH out-performs M-tree in terms of precision.

@@

@@

\caption{Linkage results on the demographic datasets}
\label{demography-quality-mtree}
\end{figure}

>>>>

Record linkage is the process of identifying records that refer to the
same real-world entities, in situations where entity identifiers are
unavailable. Records are linked on the basis of similarity between
common attributes, with every pair of records being classified as a
link or non-link depending on the degree of similarity between them.
Record linkage is usually performed in a three-step process: first
groups of similar candidate records are identified using indexing; 
pairs within the same group are then compared in more detail, and
finally classified.
%
Even state-of-the-art indexing techniques, such as Locality Sensitive
Hashing, have potential drawbacks. They may fail to group together some
true matching records with high similarity. Conversely, they may group
records with low similarity, leading to high computational overhead.
%
We propose using metric space indexing to perform \emph{complete} record
linkage, which results in a parameter-free record linkage process
combining indexing, comparison and classification into a single step
delivering complete and efficient record linkage. Our experimental
evaluation on real-world datasets from several domains shows that
linkage using metric space indexing can yield better quality than
current indexing techniques, with similar execution cost, without the
need for domain knowledge or trial and error to configure the process.

@@

To overcome data quality issues such as typographical errors and
variations (which are common in name and address values~\cite{Chr12}),
approximate string comparison functions (such as edit distance, the
Jaro-Winkler comparator, or Jaccard similarity~\cite{Chr12}) are used to
compare pairs of records, leading to a vector of similarities (one
similarity per attribute/field compared) for each pair. These similarity
vectors are then used to classify the compared record pairs into
\emph{links} (where it is assumed both records in a pair correspond to
the same real-world entity) and \emph{non-links} (where the records are
assumed to correspond to two different entities). Various
classification methods have been employed in record
linkage~\cite{Chr12,Don15}, ranging from simple threshold-based to
sophisticated clustering, supervised classification techniques, and
active learning approaches~\cite{Wan15}.

@@

Traditional record linkage systems that perform indexing prior to
comparison and classification, as illustrated in left-hand side in
Fig.~\ref{fig-rl-process}, add a further practical complexity to the
process. Indexing, comparison and classification are often conducted
using algorithms and parameters selected based on technical and
domain expertise of the user of the record linkage system, followed
by a manual assessment (auditing) of the linkage
outcomes~\cite{Chr12}. If the quality of the resulting links is not
good enough for a certain application, the linkage process needs to
be repeated with different parameter settings and possibly also
alternative algorithms. This can result in a time-consuming and
labour-intensive iterative process. A major challenge is often the
heuristic nature of indexing, where the choice of a certain indexing
technique and its parameters (including which attributes to use in indexing) will determine the final outcome of a linkage.

This paper focuses on approaches that use a similarity threshold to
classify links. Such techniques are fundamentally limited by the extent
to which true matching records are similar, and true non-matches are
dissimilar---this is dataset-dependent. Within this domain, we define
a technique to be \emph{complete} if it is guaranteed to identify all
record pairs within the specified similarity threshold. Commonly used
indexing techniques are incomplete, since they reduce computational
cost at the expense of potentially overlooking some true matches. By
definition, incomplete techniques yield lower recall than complete
techniques. Conversely, and perhaps counter-intuitively, complete
techniques can yield lower precision with some datasets. This is
discussed further in Sect.~\ref{sec-approach}.

\emph{Metric space indexing} (MSI) is a complete technique with much
lower computational overheads than a brute force approach. It allows
indexing, comparison and classification to be combined into a single
step, as illustrated in the right-hand side in
Fig.~\ref{fig-rl-process}, making the overall record linkage process
simpler, more efficient and more effective than incomplete indexing
approaches.

\smallskip
\textbf{Contribution:} The primary contribution of this paper is the
novel application of MSI to achieve complete and efficient record
linkage, without the need for complex parameter tuning. We evaluate our
approach on several real-world datasets from diverse domains and
demonstrate its advantages over existing indexing techniques for
record linkage.

@@

A different approach to indexing is the sorted neighbourhood
method~\cite{Mon96}, where the datasets to be linked are sorted
according to a \emph{sorting key} (usually a concatenation of the values
from several attributes). A sliding window is then moved over the
datasets and only records within the window are compared. Techniques
that adaptively shrink or expand the window size based on the
characteristics of the sorting key values have shown to improve both
linkage efficiency and quality~\cite{Dra12} over
%other sliding window
approaches that use fixed sizes windows.

@@

Locality sensitive hashing (LSH), originally proposed to allow efficient
nearest-neighbour search in high-dimensional spaces~\cite{Ind98}, has
been employed in record linkage as an indexing technique where attribute
values are hashed multiple times, and blocks are created from those
records that share some hash values. \emph{HARRA}~\cite{Kim10} is a
record linkage approach based on MinHash~\cite{Broder1997} and LSH
which blocks, compares, and then merges linked records in an iterative
fashion, where merged records are re-hashed to improve overall linkage
quality. \cite{Steorts2014} evaluates two LSH variations, concluding
that in order to get good results, LSH methods must be tuned to the
particular datasets being linked. This requires good quality ground
truth data which may be unavailable or expensive to obtain.

Metric space indexing (MSI) techniques provide indexing structures to
support the comparison of records in one set with those in another,
typically also offering similarity search operations. To create an MSI
data structure, it is necessary to define a distance measure between
records, with certain properties including the \emph{triangle
inequality}. Similarity search operations include
\emph{range-search(\textbf{q},d)}, where all records $\mathbf{r}$
in a database $\mathbf{D}$ within a distance $d$ of a query record
$\mathbf{q}$ are identified; \emph{nearest-neighbour(\textbf{q})},
returning the record $\mathbf{r} \in \mathbf{D}$ with smallest
distance to $\q$; and \emph{nearest-n(\textbf{q},n)}, returning
the $n$ closest records in $\mathbf{D}$ to $\mathbf{q}$.
\cite{Zezula2010} gives an overview of MSI techniques. Here we choose
one metric space indexing structure, the M-tree~\cite{paolociaccia2m},
and investigate its efficacy for record linkage.

@@

A threshold-based linkage method using R-trees~\cite{Hjaltason1998}
was described in~\cite{Li2006}. It demonstrates that high linkage
quality can be achieved using Jaccard similarity over selected record
attributes. \cite{Ciaccia97indexingmetric} shows that M-trees are
almost always more efficient than R-trees, hence their use in the
experiments described in this paper.

@@

\Cref{cora-quality} shows the precision, recall, and F-measure for
various thresholds. As expected, low thresholds give high precision and
low recall, and the reverse for high threshold values. Brute force and
M-tree give identical results. The best linkage quality, with an
F-measure of around 0.7, is achieved by a number of linkers, including
brute force, M-tree, blocking on `authors' and blocking and on all
fields, and two of the LSH configurations. All of these give similar
overall results, apart from blocking on `authors', which gives much
better quality at very high distance threshold values. This is due to
the incomplete nature of the linker, avoiding consideration of
significant numbers of high-similarity non-matches, thus avoiding these
becoming false positives and keeping precision high.

For a more detailed investigation of selected linkers, the brute force
approach is used to establish a good threshold value for the Cora
dataset. The maximum F-measure is observed at a threshold value of
70. This value is dataset-dependent; for different datasets the maximum
F-measure will occur at different thresholds. In the rest of this
section we fix the threshold value at 70.

\Cref{comparison-of-results-cora} shows greater detail for selected
linkers, showing the parameters for the experiment, the number of
distance comparisons made, the precision, recall and the F-measure
achieved by each algorithm.  In the \emph{Linker} column the algorithm
name is followed by its parameters: for LSH the number of the bands
followed by the band size, and for traditional blocking the fields used
for blocking. The number of distance comparisons is reported as a
machine-independent proxy for execution cost, since code profiling shows
that distance calculations are dominant.

@@

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-brute}
\vspace{-6mm}
\caption{Brute Force}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-mtree}
\vspace{-6mm}
\caption{M-tree}
\end{subfigure} \vspace{5mm}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-authors}
\vspace{-6mm}
\caption{Blocking on `authors' field}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-title}
\vspace{-6mm}
\caption{Blocking on `title' field}
\end{subfigure} \vspace{5mm}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-year}
\vspace{-6mm}
\caption{Blocking on `publication year' field}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-trad-combined}
\vspace{-6mm}
\caption{Blocking using the union of all fields}
\end{subfigure} \vspace{5mm}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-lsh-2-2}
\vspace{-6mm}
\caption{LSH using 2 bands of size 2}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-lsh-2-10}
\vspace{-6mm}
\caption{LSH using 2 bands of size 10}
\end{subfigure} \vspace{5mm}

\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-lsh-10-2}
\vspace{-6mm}
\caption{LSH using 10 bands of size 2}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
\includegraphics[width=\textwidth]{figures/plotLQ-cora-lsh-10-10}
\vspace{-6mm}
\caption{LSH using 10 bands of size 10}
\end{subfigure}

@@

GRAHAM STILL WRITING HERE

Birth records were linked to death records, separately for the Skye and Kilmarnock datasets, using M-tree and LSH.

It was not computationally feasible to run the brute force linker. Of the incomplete linkers, LSH was selected as it gave slightly better results for Cora.

The shingle size was set to 2 for all the LSH experiments reported, as this was found to give good results and LSH was not especially sensitive to this parameter. Results for other shingle sizes are omitted from this paper for brevity.

on the Cora dataset using all four
approaches presented in this paper: brute force, traditional blocking,
LSH and M-tree, using several selected configurations for the blocking
and LSH linkers. The distance threshold is varied between 0 and
250~\footnote{Relatively high Levenshtein distances are included since
Cora includes a number of low-similarity true matches.}. For traditional
blocking, the following Cora fields are used individually as blocking
keys: \textit{author}, \textit{title}, \textit{venue},
\textit{location}, \textit{publisher} and \textit{year}. We also use a
combined blocking key comprising all fields.

\Cref{cora-quality} shows the precision, recall, and F-measure for
various thresholds. As expected, low thresholds give high precision and
low recall, and the reverse for high threshold values. Brute force and
M-tree give identical results. The best linkage quality, with an
F-measure of around 0.7, is achieved by a number of linkers, including
brute force, M-tree, blocking on `authors' and blocking and on all
fields, and two of the LSH configurations. All of these give similar
overall results, apart from blocking on `authors', which gives much
better quality at very high distance threshold values. This is due to
the incomplete nature of the linker, avoiding consideration of
significant numbers of high-similarity non-matches, thus avoiding these
becoming false positives and keeping precision high.

\Cref{demography-quality}

\Cref{comparison-of-results-demography-skye} and \Cref{comparison-of-results-demography-kili} show the results for LSH-MinHash and M-tree on the Skye and Kilmarnock datasets.

In both cases the F1 measure achieved is better for M-tree than any of the experimental settings chosen for LSH. This is largely due to the recall for M-tree being much higher than any of the LSH experiments. For most configurations, LSH out-performs M-tree in terms of precision.

@@

\caption{Linkage results on the demographic datasets}
\label{demography-quality}
\end{figure}

@@
